Interactive effects of drought and habitat fragmentation on vital rates of an understory tropical plant

### 2021-01-29 analysis/paper/paper.Rmd
erdit spei calcs parag

The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy, emergent trees up to 55m [@rankin-de-meronaPreliminaryResultsLargescale1992], and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999]. The soils in the reserves are nutrient-poor xanthic ferrosols with poor water retention capacity despite their high clay content. Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm with a pronounced dry season from June to October (Appendix?).

@@

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot widely distributed throughout the Central Amazon [@kressDiversityDistributionHeliconia1990], and it is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010]. While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others---such as *H. acuminata*---are found at lower densities in shaded forest understories. Species in this group produce fewer infloresences and are pollinated by traplining hummingbirds that move from one plant to the next rather than defend a territory [@berryHeliconiaIdentificationGuide1991a]. In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@bruna2004]; reproductive plants will have 1-7 flowering shoots, each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds (1--3 seeds per fruit, $\bar{x}$=2 <!--# check data -->) germinate \~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999]. *Heliconia acuminata* is self-incompatible and does not reproduce vegetatively.

@@

In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *H. acuminata*. The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km. Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table <!--# insert fig ref -->). Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify surveys, during which we recorded the identity of flowering plants, the number of vegetative shoots each plant had, and the height of each plant to the tallest leaf. Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]. In 2008 we mapped the location of all established plants to the nearest 10 cm; in subsequent years we did the same with all new seedlings. There is a seven-fold range in H. acuminata density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].

@@

Precipitation and evapotranspiration data were obtained from a gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016]. In addition to precipitation records, this gridded dataset had estimates of potential evapotranspiration calculated using the Pennman-Monteith method. Monthly precipitation and potential evapotranspiration was used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010]. SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale. More negative values of SPEI indicate greater drought intensity and positive values indicate wetter than average months. SPEI between 0 and -1 is considered mild drought, between -1 and -1.5 is considered moderate drought, between -1.5 and -2 is severe drought, and less than -2 is considered extreme drought [@mckeeRelationshipDroughtFrequency1993]. A 3 month time scale for SPEI was deemed as most relevant for *H. acuminata* as it relies primarily on soil moisture and not water sources buffered over longer timescales (e.g. it lacks deep taproots to access groundwater). We calculated 3-month SPEI using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017]. We compared the SPEI values calculated using this data source to several others including near by weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017] and TRMM [@trmm3b43] and the timing of drought events is consistent among these datasets although the exact magnitude of drought differs among datasets (Appendix 1).

>>>>

The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy [@rankin-de-meronaPreliminaryResultsLargescale1992] and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999]. The soils in the reserves are nutrient-poor xanthic ferrosols; their water retention capacity is poor despite having a high clay content. Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm. There is a pronounced dry season from June to October **(Appendix?)**.

@@

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990]; it is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010]. While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others---including as *H. acuminata*---are usually found at lower densities and in the more shaded forest understory. These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991a]. In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@bruna2004]; reproductive plants have $\bar{x}$ = **X.X** flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds (1--3 seeds per fruit, $\bar{x}$=2 <!--# check data -->) germinate \~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999]. *Heliconia acuminata* is self-incompatible and does not reproduce vegetatively.

@@

In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*. The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km. Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table <!--# insert fig ref -->). Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and the identity of any flowering plants. We also surveyed plots regularly during the rainy season to identify any that flowered after the survey. Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]. In 2008 we mapped the location of all established plants to the nearest 10 cm; in subsequent years we did the same with all new seedlings. There is a seven-fold range in H. acuminata density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].

@@

Precipitation and potential evapotranspiration data were obtained from a gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016]. The estimates of potential evapotranspiration were calculated using the Pennman-Monteith method, and then along with monthly precipitation were used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010]. SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale. Positive SPEI values indicate wetter than average months<!-- instead: "...indicate the specified time period is wetter than average." because time frame doesn't have to be a month, yes?  -->, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (< -2) [@mckeeRelationshipDroughtFrequency1993]. Because *H. acuminata* lacks deep taproots to access groundwater, it probably relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales. We therefore quantified periods of drought with the 3-month SPEI, which we calculated using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017]. The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (Appendix 1).
